Canonical rings of $\bq$-divisors on Minimal Rational Surfaces

### 2015-08-08 surfaces-write-up/surface-write-up.tex
fixed some ^d$'s

For the remainder of this section we will assume $D_1, D_2, D_3,$ and $D_4$ are
distinct divisors with bi-degrees $(1,0)$
, $(1,0)$, $(0,1)$, and $(0,1)$ respectively with $D_i = V(f_i)$ for $1 \leq i \leq 4$ with $f_i \in \sco(\bida_i, \bidb_i)$.  
In order to achieve this condition on
$D_1, \ldots D_4$, it may be necessary to add in ``ghost divisors'' (i.e.
divisors with of the desired form with a coefficient $0$).
Also, note that, $f_1$ and $f_2$ are independent linear
polynomials in $u$ and $v$ and $f_3$ and $f_4$ are independent
linear polynomials in $z$ and $w$.
Analogously to Proposition ~\ref{prop:pm-span-and-basis} for the case
of $\bp^m$, all rational functions on $\hirz_m$
can be written uniquely in a form where their numerator is a function
of only $f_1,f_2,f_3$, and $f_4$.
\todo{Aaron: Do either of you think we should write out this as
a proposition more clearly? I'm fine either way, but it was something
missing which I thought I should point out.}

@@

\noindent
lying in some degree $d \in \mathbb{N}$
with $c_j \ge -\alpha_j d$ (for all $j$) satisfying $\bida_i + \sum_{j = 1}
^n \bida_j c_j = 0$ and $\bidb_i + \sum_{j=1}^n \bidb_j c_j = 0$.
\todo{Aaron: The next two paragraphs are missing a lot of $u^d$ degree prefixes. Can one of you add them in and go through the proof
to make sure there are no others missing?}
We claim such a $\beta_i$,
can be expressed as a polynomial $\beta_i(f_1,f_2,f_3,f_4)$ with 
$\beta_i(f_1,f_2,f_3,f_4) = f_i$
as elements of $\bk[\Sigma],$ which is possible as $f_1,f_2,f_3,f_4$ generate $\Sigma,$ and hence generate $\bk[\Sigma]$. Furthermore, these
generate all relations, since they allow us to reduce any monomial $\prod_{j =
1}^n f_j^{r_j}$ to the canonical form where $r_j = - \lfloor d \alpha_i
\rfloor$

 whenever $j > 4$.  To bound the degree of generation of these relations, we bound the degree of generation of the ideal $(z_i - \beta_i) \cap \ker(\psi)$ for each $i$.

We now wish to bound the degree of generation the ideal of relations generated by these.  To do this, we first fix $i$ and a relation of the form $(f_i - \beta_i)\prod_{j=1}^n f_j^{c_j}=0.$ Note that there are no relations for $i\in \{1, 2, 3, 4\}$ as $\beta_i = f_i$; thus we restrict attention to $i \ge 5$. Our goal is to find a relation that the divides this, but whose degree we can bound. We do so by considering the lattice points corresponding to the monomials appearing in the relation $(f_i - \beta_i)\prod_{j=1}^n f_j^{c_j}$, and finding some $\lambda\in \Sigma$
that we can factor out of each monomial, to get a relation below a bounded degree.  The following argument makes this precise:

@@

Consider a relation $u^d(f_i - \beta_i)\prod_{j=1}^n f_j^{c_j}=0$ with corresponding lattice point $\sigma: = (d, c_1, \ldots, c_{i-1}, c_{i}+1, c_{i+1}, \ldots, c_n) \in \Sigma$.
Then we can write $\sigma$ as a sum of $s_j e_j$'s for $j\in
\Te$ and $s_{j,k}e_{j,k}$ for $j\in \Tp, k\in \Tm$.  
For convenience, defing $d_j := \deg e_j$ (when it exists) and let the $j^{th}$
component of $e_j$ be $-\alpha_j d_j + \kappa_j$ for some $\kappa_j \in \bq$. Also, let $d_{j,k} := \deg e_{j,k}$ (when it exists) and let the $j^{th}$ component of $e_{j,k}$ be $-\alpha_j d_{j,k} + \kappa_{j,k}'$ and the $k^{th}$
component be $-\alpha_k d_{j,k} + \kappa_{j,k}''$ for $\kappa_{j,k}', \kappa_{j,k}'' \in \bq$.

For $u^d (f_i - \beta_i)\prod_{j=1}^n f_j^{c_j}$ to be a relation, it must, in particular, be an element of $R_D$. This implies that $s_i\kappa_i\ge 1$, $\sum_{j\in \Tm} s_{i,j}\kappa_{i,j}' \ge 1$, and $\sum_{j\in \Tp} s_{i,j}\kappa_{j,i}'' \ge 1$ if $i\in \Te$, $i\in \Tp$, and $i\in \Tm$ respectively.

@@

Let $F = \Psi(E) \in R_D$ and $F_\mu = \psi(E_\mu)
\in R_D$ for each $\mu \in M_i$.  
Then $F - \sum_{\mu \in M_i} F_\mu = 0$ is a relation %in degree
%\[
%	\deg(F) = \deg(E)
%\]
which divides $(z_i - \beta_i) \prod_{j = 1}^n z_j^{c_j}$. 
We can bound $\deg(F)$ (which equals $\deg(F_\mu)$ for all $\mu$) by
\[
	\deg(F) = \deg(E) \le \rho + \begin{cases}
	\ell_i \gcd(\bida_i, \bidb_i)	&\mbox{ if } i \in \Te \\
	\max_{j \in \Tm} \bigl(\ell_{i, j} (\bida_i \bidb_j - \bida_j \bidb_i)\bigr)
	&\mbox{ if } i \in \Tp \\
	\max_{j \in \Tp} \bigl(\ell_{j, i} (\bida_j \bidb_ i - \bida_i \bidb_j) \bigr)
	&\mbox{ if } i \in \Tm. \end{cases}
	\]

>>>>

For the remainder of this section we will assume $D_1, D_2, D_3,$ and $D_4$ are
distinct divisors with bi-degrees $(1,0)$
, $(1,0)$, $(0,1)$, and $(0,1)$ respectively with $D_i = V(f_i)$ for $1 \leq i \leq 4$ with $f_i \in \sco(\bida_i, \bidb_i)$.  
In order to achieve this condition on
$D_1, \ldots D_4$, it may be necessary to add in ``ghost divisors'' (i.e.
divisors with of the desired form with a coefficient $0$).
Also, note that, $f_1$ and $f_2$ are independent linear
polynomials in $u$ and $v$ and $f_3$ and $f_4$ are independent
linear polynomials in $z$ and $w$.
Analogously to Proposition ~\ref{prop:pm-span-and-basis} for the case
of $\bp^m$, all rational functions on $\hirz_m$
can be written uniquely in a form where their numerator is a function
of only $f_1,f_2,f_3$, and $f_4$.
\todo{Aaron: Do either of you think we should write out this as
a proposition more clearly? I'm fine either way, but it was something
missing which I thought I should point out. \\
Peter: The one line argument is ``Every numerator is a polynomial in $u, v, z, w$; we can generate $u$ and $v$ from $f_1$ and $f_2$, and we can generate $z$ and $w$ from $f_3$ and $f_4$.''  This can be added if you want}

@@

\noindent
lying in some degree $d \in \mathbb{N}$
with $c_j \ge -\alpha_j d$ (for all $j$) satisfying $\bida_i + \sum_{j = 1}
^n \bida_j c_j = 0$ and $\bidb_i + \sum_{j=1}^n \bidb_j c_j = 0$.
\todo{Aaron: The next two paragraphs are missing a lot of $u^d$ degree prefixes. Can one of you add them in and go through the proof
to make sure there are no others missing?\\
Peter: I added a bunch.  Should be better (although quite possibly missed a few more).}
Specifically, $\beta_i$,
is the polynomial obtained by expressing $f_i$ as a polynomial in $f_1, f_2, f_3, f_4$.
%$\beta_i(f_1,f_2,f_3,f_4)$ with 
%$\beta_i(f_1,f_2,f_3,f_4) = f_i$
%as elements of $\bk[\Sigma],$ which is possible as $f_1,f_2,f_3,f_4$ %generate $\Sigma,$ and hence generate $\bk[\Sigma]$. 
Furthermore, these
generate all relations, since they allow us to reduce any monomial $u^d\prod_{j =
1}^n f_j^{r_j}$ to the canonical form where $r_j = - \lfloor d \alpha_i
\rfloor$
whenever $j > 4$.  To bound the degree of generation of these relations, we bound the degree of generation of the ideal $(z_i - \beta_i) \cap \ker(\psi)$ for each $i$.

We now wish to bound the degree of generation the ideal of relations generated by these.  To do this, we first fix $i$ and a relation of the form $u^d(f_i - \beta_i)\prod_{j=1}^n f_j^{c_j}=0.$ Note that there are no relations for $i\in \{1, 2, 3, 4\}$ as $\beta_i = f_i$; thus we restrict attention to $i \ge 5$. Our goal is to find a relation that the divides this, but whose degree we can bound. We do so by considering the lattice points corresponding to the monomials appearing in the relation $u^d(f_i - \beta_i)\prod_{j=1}^n f_j^{c_j}$, and finding some $\lambda\in \Sigma$
that we can factor out of each monomial, to get a relation below a bounded degree.  The following argument makes this precise:

@@

Consider a relation $u^d(f_i - \beta_i)\prod_{j=1}^n f_j^{c_j}=0$ and let it correspond to the lattice point $\sigma: = (d, c_1, \ldots, c_{i-1}, c_{i}+1, c_{i+1}, \ldots, c_n) \in \Sigma$.
Then we can write $\sigma$ as a sum of $s_j e_j$'s for $j\in
\Te$ and $s_{j,k}e_{j,k}$ for $j\in \Tp, k\in \Tm$.  
For convenience, define $d_j := \deg e_j$ (when it exists) and let the $j^{th}$
component of $e_j$ be $-\alpha_j d_j + \kappa_j$ for some $\kappa_j \in \bq$. Also, let $d_{j,k} := \deg e_{j,k}$ (when it exists) and let the $j^{th}$ component of $e_{j,k}$ be $-\alpha_j d_{j,k} + \kappa_{j,k}'$ and the $k^{th}$
component be $-\alpha_k d_{j,k} + \kappa_{j,k}''$ for $\kappa_{j,k}', \kappa_{j,k}'' \in \bq$.

Since $u^d (f_i - \beta_i)\prod_{j=1}^n f_j^{c_j}$ is a relation, each monomial of it must be an element of $R_D$. This implies that $s_i\kappa_i\ge 1$, $\sum_{j\in \Tm} s_{i,j}\kappa_{i,j}' \ge 1$, and $\sum_{j\in \Tp} s_{i,j}\kappa_{j,i}'' \ge 1$ if $i\in \Te$, $i\in \Tp$, and $i\in \Tm$ respectively.

@@

Let $F = \Psi(E) \in R_D$ and $F_\mu = \psi(E_\mu)
\in R_D$ for each $\mu \in M_i$.  
Then $F - \sum_{\mu \in M_i} F_\mu = 0$ is a relation %in degree
%\[
%	\deg(F) = \deg(E)
%\]
which divides $u^d(z_i - \beta_i) \prod_{j = 1}^n z_j^{c_j}$. 
We can bound $\deg(F)$ (which equals $\deg(F_\mu)$ for all $\mu$) by
\[
	\deg(F) = \deg(E) \le \rho + \begin{cases}
	\ell_i \gcd(\bida_i, \bidb_i)	&\mbox{ if } i \in \Te \\
	\max_{j \in \Tm} \bigl(\ell_{i, j} (\bida_i \bidb_j - \bida_j \bidb_i)\bigr)
	&\mbox{ if } i \in \Tp \\
	\max_{j \in \Tp} \bigl(\ell_{j, i} (\bida_j \bidb_ i - \bida_i \bidb_j) \bigr)
	&\mbox{ if } i \in \Tm. \end{cases}
	\]
